---
title: "blue_bike_eda"
output:
  pdf_document: default
  html_document: default
date: "2024-10-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r including_packages, include=FALSE}
#including packages
library(dplyr)
library(purrr)
library(readr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(plotly)
```

```{r load_files, include=FALSE}
setwd('/Users/mirayagupta/Desktop/DS104/bb2023')
csv_file_list = dir()
csv_file_list
df_list = map(csv_file_list, read_csv)
data = bind_rows(df_list)
```


```{r}
dim(data)
colnames(data)
summary(data)
```
```{r}
print('start station id')
sum(is.na(data$`start station id`))
sum(is.na(data$start_station_id))
na_start_indices1 = which(is.na(data$`start station id`))
na_start_indices2 = which(is.na(data$start_station_id))
unique(na_start_indices1 == na_start_indices2)
# seems like there are certain indices for which there is no station start id available
print('start station name')
sum(is.na(data$`start station name`))
sum(is.na(data$start_station_name))
# for most of these indices, the station name is available. Not available for 1060 rides. This can be imputed using latitude/longitude

print('end station id')
sum(is.na(data$`end station id`))
sum(is.na(data$end_station_id))
na_end_indices1 = which(is.na(data$`end station id`))
na_end_indices2 = which(is.na(data$end_station_id))
unique(na_end_indices1 == na_end_indices2)
# seems like there are certain indices for which there is no end station id available
print('end station name')
sum(is.na(data$`end station name`))
sum(is.na(data$end_station_name))
# for most of these indices, the station name is available. Not available for 21564 rides. This can be imputed using latitude/longitude

#data = data %>% select(!c(started_at, ended_at, `start station id`, `start station name`, `start station latitude`, 
      #             `start station longitude`, `end station id`, `end station name`, `end station latitude`, `end station longitude`))

colnames(data)

for (column in colnames(data)) {
  print(column)
  print(sum(is.na(data[[column]])))
}
```

```{r clean_columns}
#coalescing repeated columns

data = data %>%
  mutate(starttime = coalesce(starttime, started_at)) %>%
  mutate(stoptime = coalesce(stoptime, ended_at)) %>%
  mutate(start_station_id = coalesce(`start station id`, as.numeric(start_station_id))) %>%
  mutate(start_station_name = coalesce(`start station name`, start_station_name)) %>%
  mutate(start_lat = coalesce(`start station latitude`, start_lat)) %>%
  mutate(start_lng = coalesce(`start station longitude`, start_lng)) %>%
  mutate(end_station_id = coalesce(`end station id`, as.numeric(end_station_id))) %>%
  mutate(end_station_name = coalesce(`end station name`, end_station_name)) %>%
  mutate(end_lat = coalesce(`end station latitude`, end_lat)) %>%
  mutate(end_lng = coalesce(`end station longitude`, end_lng)) %>%
  mutate(usertype = coalesce(member_casual, usertype))
```
```{r}
sum(is.na(data$starttime))
sum(is.na(data$bikeid))
```
Since there are this many NA values in the bikeid variable (due to a change in how data was collected in the middle of 2023) it is not possible to calculate utilisation at an individual bike level for 2023. 

An alternative I will use is to try and look at the number of bikes in use per station. 

```{r dropping_columns}
data = data %>% select(-bikeid, -started_at, -ended_at,
                       -`start station id`, -`start station name`, 
                       -`end station id`, -`end station name`,
                       -`start station latitude`, -`start station longitude`,
                       -`end station latitude`, -`end station longitude`,
                       -`postal code`, -member_casual)
colnames(data)
```

```{r plotting_data}
sample = sample_n(data, 1000)
par(mfrow = c(3, 3))
# Assuming sample$starttime is your POSIXt data
hist(sample$starttime, breaks = "hours", main = "Histogram of Start Times", xlab = "Start Time", col = "lightblue")
# Assuming sample$stoptime is a POSIXt object
hist(sample$stoptime, breaks = "hours", main = "Histogram of Stop Times", xlab = "Stop Time", col = "lightgreen")
barplot(table(sample$usertype))
barplot(table(sample$rideable_type))
hist(sample$start_lng)
hist(sample$end_lng)
hist(sample$start_lat)
hist(sample$end_lat)
```

```{r laoding_stationdata, include=FALSE}
setwd('/Users/mirayagupta/Desktop/DS104')
station_data = read_csv('current_bluebikes_stations.csv')
```


```{r cleaning_stationdata}
new_headers = station_data[1,]
colnames(station_data) = new_headers
station_data = station_data[-1, ]
colnames(station_data)
station_data = station_data %>%
                  mutate(`Total Docks` = as.numeric(`Total Docks`)) #this is coded as a character
dim(station_data)
summary(station_data)
```


```{r merging_datasets}

# merge for the start station
data = data %>%
  left_join(station_data, by = c("start_station_name" = "NAME")) 

print(colnames(data))

data = data %>%
  rename(start_station_number = `Number`,
         start_station_municipality = `Municipality`,
         start_seasonal_status = `Seasonal Status`, 
         start_station_docks = `Total Docks`, 
         start_station_id_ = `Station ID (to match to historic system data)`)

# merge for the end station
data = data %>%
  left_join(station_data, by = c("end_station_name" = "NAME"))

print(colnames(data))

data = data %>%
  rename(end_station_number = `Number`,
         end_station_municipality = `Municipality`,
         end_seasonal_status = `Seasonal Status`, 
         end_station_docks = `Total Docks`, 
         end_station_id_ = `Station ID (to match to historic system data)`)

# start_station_id_ represents the station ID from the station data as opposed to the ride data
# end_station_id_ represents the station ID from the station data as opposed to the ride data
```
```{r}
#Adding information for month and day of week
data$month = month(data$starttime)
data$day_of_week = wday(data$starttime)
```

```{r visualising_docks}
docks_data = station_data %>%
  select(`NAME`, `Total Docks`)
docks_data = docks_data %>%
                arrange(desc(`Total Docks`))

top_docks_data = head(docks_data, 10)

par(mfrow = c(1, 2)) #this doesn't work with ggplot graphics so I have to find another way to arrange them

ggplot(data=top_docks_data, aes(x=reorder(NAME, -`Total Docks`), y=`Total Docks`)) +
  geom_bar(stat="identity") + 
  coord_flip() + 
  labs(title = "Number of Docks (Top 10)", x='', y='') + 
  theme_minimal()

#tried to bring these plots on the same axes but base R is not good with that 
hist(docks_data$`Total Docks`, main='Distribution of Number of Docks', xlab='', ylab='')
boxplot(docks_data$`Total Docks`, main='Distribution of Number of Docks', horizontal=TRUE)
```
The distribution of the number of docks is heavily right-skewed. 
```{r}
#Grouping rides by start station
#managing NAs in start station name/id
sum(is.na(data$start_station_name)) #8
sum(is.na(data$start_station_id)) #3200875
sum(is.na(data$start_station_id_)) #24124

data_by_start_station = data %>%
                          group_by(start_station_name) %>%
                          summarize(count = n())

sum(is.na(docks_data$`Total Docks`))

data_by_start_station = data_by_start_station %>%
  left_join(docks_data, by = c("start_station_name" = "NAME")) 

data_by_start_station = data_by_start_station %>%
                          arrange(desc(`Total Docks`))

dim(data_by_start_station)

data_by_start_station$rides_by_docks = round(data_by_start_station$count/data_by_start_station$`Total Docks`)

data_by_start_station = data_by_start_station %>%
                          arrange(desc(rides_by_docks))

data_by_start_station_top = head(data_by_start_station, 10)

ggplot(data=data_by_start_station_top, aes(x=reorder(start_station_name, -rides_by_docks), y=`rides_by_docks`)) +
  geom_bar(stat="identity") + 
  coord_flip() + 
  labs(title = "Ratio of Rides to Docks (Top 10)", x='', y='') + 
  theme_minimal()

data_by_start_station = data_by_start_station %>%
                          arrange(desc(count))

data_by_start_station_top = head(data_by_start_station, 10)

ggplot(data=data_by_start_station_top, aes(x=reorder(start_station_name, -count), y=count)) +
  geom_bar(stat="identity") + 
  coord_flip() + 
  labs(title = "Total Rides (Top 10)", x='', y='') + 
  theme_minimal()

plot(data_by_start_station$`Total Docks`, data_by_start_station$count, xlab='Number of Docks at Station', ylab='Number of Rides from Station')
```
It appears that there is lots of variation in the number of rides taken from a station regardless of how many docks it has. This indicates that the total number of docks is not a good linear predictor for the number of rides from a station. This could means that there is a mismatch between demand and supply and there is room to optimise the number of docks at each station. 


```{r}
data_by_start_station = data %>%
                          group_by(start_station_name) %>%
                          summarize(count = n())
```